Comparison of Several Stochastic and Deterministic Derivative-free Global Optimization Algorithms

### 2019-06-26 2019/MOTOR_Sovrasov/slides/slides.tex
MOTOR: updae slides

\begin{frame}
  \frametitle{Methods}
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item Have complicated internal structure in multidimensional case;
      \item Usually store and use the whole history of trials accumulated during search;
      \item Require non-trivial parallelization schemes;
      \item Under some assumptions convergence to the global solution is guaranteed.
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Have relative simple internal structure;
      \item Require constant amount of memory to store internal state of some random process or individuals of population;
      \item In most cases allows wide-scale parallelization by trials or by running many copies of the same method
      with different seeds in parallel;
      \item Convergence is guaranteed in probabilistic sense only.
    \end{itemize}
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Results of sequential methods}

  \begin{table}
  \begin{center}
  \caption{Averaged number of trials executed by sequential methods for solving the test
  optimization problems}
  \resizebox{\textwidth}{!}{%
    \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
      \hline
      & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
    \hline
    \(F_{GR}\)     & 193.1 &  \textbf{158.3} & 400.3 & 182.3 & 214.9 & 947.2 & 691.2 & 1257.3 &
  1336.8 \\
    \hline
    GKLS 2d Simple &  254.9 & 217.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2
  & 1251.5 \\
    \hline
    GKLS 2d Hard   &  728.7 & \textbf{488.0} & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 &
  1041.1 & 2532.2 \\
    \hline
    GKLS 3d Simple &  1372.1 & 1195.3 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 &
  5956.94 & 3856.1 \\
    \hline
    GKLS 3d Hard   &  3636.1 & \textbf{1930.5} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 &
  6914.3 & 7843.2 \\
    \hline
    GKLS 4d Simple &  26654.1 & 11095.7 & 37436.8 & \textbf{7824.3} & 15994.1 & 41514.3 &
  21417.9 & 19157.7 & 59895.4 \\
    \hline
    GKLS 4d Hard   &  54536.8 &  \textbf{23167.8} &  73779.3 &  23204.4 &  54489.9 &  80247.2
  &  68815.5 &  27466.1 &  109328.1  \\
    \hline
    GKLS 5d Simple &  29810.0 & 11529.0 & 143575.0 & \textbf{7166.5} & 13970.5 & 52647.6 &
  34255.3 & 73074.5 & 91580.4 \\
    \hline
    GKLS 5d Hard   &  113129.1 & 67652.7 & 165192.8 & \textbf{66327.4} & 164390.6 & 138766.2
  & 116973.1 & 105496.9 & 155123.8 \\
    \hline
  \end{tabular}}
    \label{tab:trials}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of sequential methods}
  \begin{table}
  \begin{center}
  \caption{Number of test optimization problems solved by sequential methods}
    \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
      \hline
      & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
    \hline
    \(F_{GR}\)     &  100 & 100 & 76  & 100 & 100 & 97  & 96  & 96  & 67\\
    \hline
    GKLS 2d Simple &  100 & 100 & 85  & 100 & 100 & 100 & 100 & 98  & 90\\
    \hline
    GKLS 2d Hard   &  100 & 100 & 74  & 100 & 100 & 100 & 93  & 85  & 77 \\
    \hline
    GKLS 3d Simple &  100 & 97  & 75  & 100 & 100 & 100 & 89  & 86  & 44 \\
    \hline
    GKLS 3d Hard   &  100  & 99   & 72   & 100  & 99   & 100  & 88   & 77   & 43 \\
    \hline
    GKLS 4d Simple &  100 & 100 & 46  & 100 & 100 & 94  & 78  & 59  & 16 \\
    \hline
    GKLS 4d Hard   &  100 & 100 & 47  & 99  & 97  & 94  & 72  & 32  & 10  \\
    \hline
    GKLS 5d Simple &  100 & 100 & 68  & 100 & 100 & 98  & 100 & 77  & 9  \\
    \hline
    GKLS 5d Hard   &  97  & 99  & 42  & 100 & 90  & 79  & 84  & 48  & 8 \\
    \hline
    \end{tabular}
    \label{tab:solved}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Parallel optimization methods}
  \begin{itemize}
    \item Parallel AGS (Globalizer system);
    \item Parallel ACO (MIDACO system).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item AGS\(l\) method implemented in the Globalizer system has demonstrated the convergence
    speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
    implementations of which are available;
    \item the stochastic optimization methods inferior to the deterministic ones in the convergence
  speed and in reliability. It is manifested especially strongly on more complex multiextremal
  problems;
    \item the parallel version of the Globalizer system demonstrates good speedup
  when running on several nodes, on each of which a single objective function value per iterations is
  computed. When solving the problems with fast computable objective functions and using multiple
  threads on the nodes, the speedup for the Globalizer system degrade with increasing
  number of nodes;
    \item the MIDACO system is the most suitable for simple global optimization problems with the
  fast-computable objective functions. In this case, MIDACO is reliable enough and provides a linear
  speedup with increasing number of nodes and threads executed on these ones in parallel.
    \end{itemize}
    %Future work:
    %\begin{itemize}
    %  \item
    %\end{itemize}
\end{frame}
{
\unnumbered
\begin{frame}{{}}
  \frametitle{Q\&A}
  \begin{center}
    \Large{Contacts:}
\vspace{0.5cm}

>>>>

\begin{frame}
  \frametitle{Methods}
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item Have complicated internal structure in multidimensional case;
      \item Usually store and use the whole history of trials accumulated during search;
      \item Under some assumptions convergence to the global solution is guaranteed.
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Have relative simple internal structure;
      \item Require constant amount of memory to store internal state of some random process or individuals of population;
      \item Convergence is guaranteed in probabilistic sense only.
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Hyperparameters tuning in AGS}
  AGS selects the best intervals according to the following characteristic:
  \begin{displaymath}
    R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{(rH)^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{rH},1<i<k+1
  \end{displaymath}
  where $L$ is current estimation on Holder constant, $r$ is reliability coefficient.
  AGS is very sensitive to choice of $r$:
  \begin{itemize}
    \item low value can lead to fast convergence to a local minima
    \item high value leads to very slow convergence
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Hyperparameters tuning in AGS}
  In order to resolve the problem of choosing $r$ to some extent,
  let us use the following scheme:
  \begin{itemize}
    \item execute $q$ iterations of AGS with $r=r_{max}$;
    \item execute $q$ iterations of AGS with $r=r_{min}$;
    \item repeat the above steps either until convergence or until the allowed number of iterations are
  exhausted.
\end{itemize}
  Now we have 3 parameters instead of one, but the method is not so sensitive to them.
  Let $q=50\cdot\log_2(N-1)\cdot N^2$, $r_{min}=3,\:r_{max}=2\cdot r_{min}$ in all the future experiments.
\end{frame}

@@

\begin{frame}
  \frametitle{Experimental setup}
  \begin{table}
  \begin{center}
  \caption{Trials limits and relative precision for the test problem classes}
    \begin{tabular}{|l|{c}|{c}|}
      \hline
    Problems class & Trials limit & $\alpha$\\
    \hline
    \(F_{GR}\) & 5000 & 0.01 \\
    \hline
    GKLS 2d Simple & 8000 & 0.01 \\
    \hline
    GKLS 2d Hard & 9000 & 0.01 \\
    \hline
    GKLS 3d Simple & 15000 & 0.01 \\
    \hline
    GKLS 3d Hard & 25000 & 0.01 \\
    \hline
    GKLS 4d Simple & 150000 & $\sqrt[4]{10^{-6}}$ \\
    \hline
    GKLS 4d Hard & 250000 & $\sqrt[4]{10^{-6}}$ \\
    \hline
    GKLS 5d Simple & 350000 & $\sqrt[5]{10^{-7}}$ \\
    \hline
    GKLS 5d Hard & 600000 & $\sqrt[5]{10^{-7}}$ \\
    \hline
    \end{tabular}
    \label{tab:limits}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Experimental setup}
  \begin{table}
  \begin{center}
  \caption{Class-specific parameters of the optimization algorithms}
    \begin{tabular}{|l|{c}|{c}|{c}|}
      \hline
      & AGS & CRS & DE\\
    \hline
    \(F_{GR}\) & \(r=3\) & popsize=150 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Simple & \(r=4.6\) & popsize=200 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Hard & \(r=6.5\) & popsize=400 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 3d Simple & \(r=3.7\) & popsize=1000 & mutation=(1.1,1.9), popsize=70 \\
    \hline
    GKLS 3d Hard & \(r=4.4\) & popsize=2000 & mutation=(1.1,1.9), popsize=80 \\
    \hline
    GKLS 4d Simple & \(r=4.7\) & popsize=8000 & mutation=(1.1,1.9), popsize=90 \\
    \hline
    GKLS 4d Hard & \(r=4.9\) & popsize=16000 & mutation=(1.1,1.9), popsize=100 \\
    \hline
    GKLS 5d Simple & \(r=4\) & popsize=25000 & mutation=(1.1,1.9), popsize=120 \\
    \hline
    GKLS 5d Hard & \(r=4\) & popsize=30000 & mutation=(1.1,1.9), popsize=140 \\
    \hline
  \end{tabular}
    \label{tab:params}
  \end{center}
  \end{table}
  \begin{itemize}
    \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
    \item in the SDA method, the parameter \(visit=2.72\).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Results of different methods}

  \begin{table}
  \begin{center}
  \caption{Averaged number of trials executed by the methods for solving the test
  optimization problems}
  \resizebox{\textwidth}{!}{%
    \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
      \hline
      & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
    \hline
    \(F_{GR}\)     & 193.1 & 248.3 & 400.3 & \textbf{182.2} & 214.9 & 947.2 & 691.2 & 1257.3 & 1336.8 \\
    \hline
    GKLS 2d Simple & 254.9 & 221.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2 & 1251.5 \\
    \hline
    GKLS 2d Hard   & \textbf{728.7} & 785.0 & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 & 1041.1 & 2532.2 \\
    \hline
    GKLS 3d Simple &  1372.1 & 1169.5 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 & 5956.9 & 3856.1 \\
    \hline
    GKLS 3d Hard   &  3636.1 & \textbf{1952.1} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 & 6914.3 & 7843.2 \\
    \hline
    GKLS 4d Simple &  5729.8 & \textbf{4919.1} & 19883.6 & 7328.8 & 15010.0 & 41484.8 & 22066.0 & 6271.2 & 29359.2 \\
    \hline
    GKLS 4d Hard   &  13113.4 & \textbf{12860.1} & 27137.4 & 22884.4 & 55596.1 & 80220.1 & 68048.0 & 12487.6 & 58925.5  \\
    \hline
    GKLS 5d Simple &  \textbf{5821.5} & 6241.3 & 62921.7 & 5966.1 & 10795.5 & 52609.2 & 34208.8 & 20859.4 & 69206.8 \\
    \hline
    GKLS 5d Hard   &  \textbf{17008.6} & 21555.1 & 87563.9 & 61657.3 & 148637.8 & 138011.8 & 115634.6 & 26850.0 & 141886.5 \\
    \hline
  \end{tabular}}
    \label{tab:trials}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of different methods}
  \begin{table}
  \begin{center}
  \caption{Number of test optimization problems solved by the methods}
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     &  100 & 100 & 76 & 100 & 100 & 97 & 96 & 96 & 67\\
  \hline
  GKLS 2d Simple &  100 & 100 & 85 & 100 & 100 & 100 & 100 & 98 & 90\\
  \hline
  GKLS 2d Hard   &  100 & 97 & 74 & 100 & 100 & 100 & 93 & 85 & 77 \\
  \hline
  GKLS 3d Simple &  100 & 100 & 75 & 100 & 100 & 100 & 89 & 86 & 44 \\
  \hline
  GKLS 3d Hard   &  100 & 100 & 72 & 100 & 99 & 100 & 88 & 77 & 43 \\
  \hline
  GKLS 4d Simple &  100 & 100 & 74 & 100 & 100 & 94 & 82 & 68 & 72 \\
  \hline
  GKLS 4d Hard   &  100 & 100 & 60 & 99 & 99 & 94 & 73 & 55 & 69  \\
  \hline
  GKLS 5d Simple &  100 & 100 & 86 & 100 & 100 & 98 & 100 & 88 & 82  \\
  \hline
  GKLS 5d Hard   &  100 & 100 & 77 & 100 & 93 & 79 & 86 & 77 & 78 \\
  \hline
  \end{tabular}
    \label{tab:solved}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of different methods}
  \begin{figure}[ht]
    \centering
    \subfloat[4d Simple]{{\includegraphics[width=.5\textwidth]{gklss4d.pdf}}\label{fig:s4d}}
    \subfloat[4d Hard]{{\includegraphics[width=.5\textwidth]{gklsh4d.pdf}}\label{fig:h4d}}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Results of different methods}
  \begin{figure}[ht]
    \centering
    \subfloat[5d Simple]{{\includegraphics[width=.5\textwidth]{gklss5d.pdf}}\label{fig:s4d}}
    \subfloat[5d Hard]{{\includegraphics[width=.5\textwidth]{gklsh5d.pdf}}\label{fig:h4d}}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Robustness of the AGS-AR to hyperparemeters}
  \begin{figure}[ht]
    \centering
    \includegraphics[width=.6\textwidth]{ar_stab.pdf}
    \caption{Operating characteristics of AGS and AGS-AR with different hyperparameters
    when solving problems from the GKLS 5d Simple classe.}
    \label{fig:stability}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item The proposed modification of the stock AGS, AGS-AR allows to pay less attention to initial hyperparameter tuning and
      performs on-par with properly tuned AGS;
      \item AGS-AR method has demonstrated the convergence
      speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
      implementations of which are available in web;
      \item the stochastic optimization methods inferior to the deterministic ones in the convergence
    speed and in reliability. It is manifested especially strongly on more complex multiextremal
    problems.
    \end{itemize}
\end{frame}

{
\unnumbered
\begin{frame}{{}}
  \frametitle{Q\&A}
  \begin{center}
    \Large{Contacts:}
\vspace{0.5cm}
